---
title: "R quickstart"
author: "Peter Solymos"
output: html_document
---

## Basic data structures and operations in R

R is a great calculator:

```{r}
1 + 2
```

Assign a value and print an object:

```{r}
print(x)
(x = 2) # shorthand for print
x == 2 # logical operator, not assignment
y <- x + 0.5
y # another way to print
```

Logical operators:

```{r}
x == y # equal
x != y # not eaqual
x < y # smaller than
x >= y # greater than or equal
```

Vectors and sequences:

```{r}
x <- c(1, 2, 3)
x
1:3
seq(1, 3, by = 1)

rep(1, 5)
rep(1:2, 5)
rep(1:2, each = 5)
```

Vector operations, recycling:

```{r}
x + 0.5
x * c(10, 11, 12, 13)
```

Indexing vectors, ordering:

```{r}
x[1]
x[c(1, 1, 1)] # a wey of repeatig values
x[1:2]
x[x != 2]
x[x == 2]
x[x > 1 & x < 3]
order(x, decreasing=TRUE)
x[order(x, decreasing=TRUE)]
rev(x) # reverse
```

Character vectors, `NA` values, and sorting:

```{r}
z <- c("b", "a", "c", NA)
z[z == "a"]
z[!is.na(z) & z == "a"]
z[is.na(z) | z == "a"]
is.na(z)
which(is.na(z))
sort(z)
sort(z, na.last=TRUE)
```

Matrices and arrays:

```{r}
(m <- matrix(1:12, 4, 3))
matrix(1:12, 4, 3, byrow=TRUE)

array(1:12, c(2, 2, 3))
```

Attribues:

```{r}
dim(m)
dim(m) <- NULL
m
dim(m) <- c(4, 3)
m
dimnames(m) <- list(letters[1:4], LETTERS[1:3])
m
attributes(m)
```

Matrix indices:

```{r}
m[1:2,]
m[1,2]
m[,2]
m[,2,drop=FALSE]
m[2]

m[rownames(m) == "c",]
m[rownames(m) != "c",]
m[rownames(m) %in% c("a", "c", "e"),]
m[!(rownames(m) %in% c("a", "c", "e")),]
```

Lists and indexing:

```{r}
l <- list(m = m, x = x, z = z)
l
l$ddd <- sqrt(l$x)
l[2:3]
l[["ddd"]]
```

Data frames:

```{r}
d <- data.frame(x = x, sqrt_x = sqrt(x))
d
```

Structure:

```{r}
str(x)
str(z)
str(m)
str(l)
str(d)
str(as.data.frame(m))
str(as.list(d))
```

Summary:

```{r}
summary(x)
summary(z)
summary(m)
summary(l)
summary(d)
```

> ### Key concepts
> 
> * a matrix is a vector with `dim` attribute, elements are in same mode,
> * a data frame is a list where length of elements match and elements can be in different mode.



Up next: Tidyverse, ggplot
